---
title: "Introduction to R | Data"
theme : united
author: "EcoHealth Alliance"
output: github_document
highlight: tango
---
<img src = "rlogo.png" style="position:absolute;top:0px;left:0px;"/>

## Contents
[Dataframes](#df)

[Dataframe structure](#str)

[Numerics & Integers](#numint)

[Factors & Characters](#char)

[Vectors](#vec)

## <a name = "df"><a/>Dataframes 

We have learned how to load dataframes. Now we will explore the general structure 
of our data and what types of variables it contains. 

The first row in a dataframe is called the *header* and holds all column (variable) names. 
The first column of the dataframe usually consists of row names. However, row names aren't 
always present in dataframes. All rows after the *header* are called *data rows*. 
Each member of a row is called a *cell*. 

We will start by loading the *cgd* (Chronic Granulotomous Disease data) dataset built into the *survival* package.package.
```{r, echo = FALSE}
# First we need to load the package
library(survival)

# Then load the cgd dataframe
data("cgd") 
```

We can quickly look at the number of rows in our dataframe
```{r}
nrow(cgd)    # Number of rows in cgd
```

Likewise we can look at the number of columns, or in this case variables
```{r}
ncol(cgd) # Number of variables in cgd
```

If we want to get a general idea of what our dataframe looks like, we can take a look at its first (head) entries
```{r}
head(cgd) # First 6 elements
```

or last (tail) ones
```{r}
tail(cgd) # Last 6 elements
```

We can also determine the value of a particular cell by specifying the row and column number from which we wish to extract values.

```{r}
cgd[2,3]   # Value from second row and third column
```

Additionally, we can find out how many elements are within a single column of the dataframe

```{r}
length(cgd$center) #should be the same as nrow
```

### EXERCISE 1

1. Load the in-built *iris* dataframe
2. How many rows are in this dataframe? 
3. Get the first 6 elements of the dataframe

## <a name = "str"><a/> Dataframe structure

We can also check the structure of our dataframe

```{r}
#Visualize the structure of the dataframe using the str() function

str(cgd)
```

We can now visualize the number of observations (203) and variables (16) in this dataframe. Likewise, we can see the first elements for each variable (column) and the data class to which it belongs to. In this case we have 4 types of data *integers* (e.g. id, age), *numerics* (e.g. height, weight), *factors* (e.g. treatment, sex) and *date*. 

###EXERCISE 2

1. Load the built-in *iris* dataframe using the ```data()``` function.
2. Check the structure of this dataframe
3. What types of variables does it contain? 

We will now talk about the different types of data R deals with

***
## <a name = "numint"><a/> Numeric & integer variables

R understands *numerics* as numbers with decimals. For example, let's extract the *weight* vector from the *cgd* dataframe. 

```{r}
weight = cgd$weight

class(weight) #The class function tells us what data type we are dealing with

```

When using the ```class()``` function, we can observe that *weight* is indeed a numeric type variable. However, what happens when we do the same with the *age* vector?

```{r}
age = cgd$age
class(age)
```

We see that even though *age* is also a number, it is classified as a *integer* variable. Non-decimal variables are called *integers*.

We can transform these integer values into numeric (with decimals) by calling the ```as.numeric()``` function.

```{r}
age_num = as.numeric(age)
class(age_num)
```

We can also coerce numeric values into integer variables by using the ```as.integer()``` function.

```{r}
weight_int = as.integer(weight)  # Transform from numeric to integer

class(weight_int)                # Get new integer vector
```


Note that R coerces numeric variables into integers by deleting any decimals or *rounding down*.
```{r}
print(weight_int)
```

Keep in mind that the default for R is to interpret numbers as numerics. So when you import a dataframe from another source integer values will be assumed to be numeric, that is having decimal values. 

### EXERCISE 3

1. Extract the *Petal.Length* column into an independent vector called *PetalLen*
2. Determine what data type this new vector is using the ```class()``` function.
3. Coerce this vector into an integer

***

## <a name = "char"><a/> Factors & Characters

Factors are categorical variables, meaning that they can only take a limited number of different values. For example, in the *sex* column of the *cgd * dataframe, we only observe 2 possible categories, male and female. 

```{r}
class(cgd$sex)
levels(cgd$sex)
```

Though dealing with a limited categories is useful, however as the number of these category levels increase it may become unmanageable. In that case we may prefer working with character variables. These are string (or letter) values, however they are not restricted to a limited number of categories. Like factors, Character strings are also enclosed in quotations (```" "```).

We can easily convert factors into characters with the ```as.character()``` function. Let's do it for the *center* column within the *cgd* dataframe:

```{r}
cgd$center = as.character(cgd$center)

class(cgd$center)
```
Note that by using the **$** operator we are not creating a new vector as we had done previously, rather we are choosing the column (center) within the dataframe (cgd) and applying these changes directly to the dataframe. Such that:

```{r}
str(cgd)
```

We can see that that the class for *center* has changed.

### EXERCISE 4

1. Transform the *Species* column into a character vector. 

