---
title: "Importing and Exporting data files"
author: "EcoHealth Alliance"
date: "March 2, 2016"
output: html_document
---

<img src = "rlogo.png" style="position:absolute;top:0px;left:0px;"/>

We might want to import data stored in other formats (e.g. CSV, SAS, Excel) for analysis in R. 

### CSV files 
To import csv files we can use the *read.csv*() function.

``` {r}
data = read.csv("MyData.csv", header= TRUE, sep = ",")
```
Note that ```header = TRUE``` is indicating that the first row of the dataframe should not be considered as part of the data since it holds variable names. If this is not the case for your data then you would set ```header = FALSE``` within the function.

The *sep* parameter indicates the delimiter that should be considered when separating your data into columns. Since this is a **Comma separated value** file, the delimter we set ```sep = "," ```.

### TXT files

Flat text files can be imported into R using *read.table*() 

```{r} 
data = read.table("MyData.txt, header = TRUE, sep = "\t")
```
*read.table*() is a generalized version of *read.csv*() and can also be used to import CSV files. In this case data was tabulated so we set ```sep = "\t"``` as the delimiter. 

### SPSS 
R can import SPSS data sets with the *read.spss*() function from the *foreign* package. 

```{r}
# Load the foreign package
librar(foreign)
data = read.spss("MyData.sav", use.value.label = TRUE, to.data.frame = TRUE)
```
```use.value.label = TRUE``` converts all variables with labels in SPSS into factors in R, while ```to.data.frame = TRUE``` converts the data into an R dataframe. 

Alternatively, we can use the *spss.get*() function from the *Hmisc* package. 

```{r}
install.packages("Hmisc") #if not previously installed
library(Hmisc)
data = spss.get("MyData.sav", use.value.label = TRUE)
```
Keep in midne that *foreign* is a defaultlibrary to R, *Hmisc* must be installed before loading.

### Stata

To import Stata files we will also use the *foreign* package and its *read.dta*() function.
```{r} 
data = read.dta("MyData.dta")
```
### SAS

There are different ways in which you can import SAS files into R, however the most convinient is to export your SAS dataset into CSV and then import it using the *read.csv*() function as mentioned previously. 

Another way you can import your SAS data is by using the *sasxport.get*() function from the *Hmisc* library

```{r}
data = sasxport.get("MyData.xpt")
```
