g---
title: "Introduction to R | Data Manipulation"
author: "EcoHealth Alliance"
date: "February 1, 2016"
output: html_document
---

<img src = "rlogo.png" style="position:absolute;top:0px;left:0px;"/>

## Contents

[Dataframes](#df)

[Loading dataframes](#load)

[Drop or add columns](#drop)

[Data summary](#dsum)

[Change column names](#colname)

[Subsetting the dataframe](#subset)


## <a name = "df"><a/>Dataframes 

Data can be stored in tables as dataframe objects. We can create dataframes by merging different vectors of the same length with the data.frame() function. 

```{r}
numbers = c(1,2,3)                         # Create 3 vectors of equal length

letters = c("a", "b", "c")

logical = c(TRUE, TRUE, FALSE)

df = data.frame(numbers, letters, logical) # Create data-frame

df                                         # Print data-frame                     
```

There are a few dataframes built-in to R that can be loaded by calling upon them. For example the *mtcars* dataset, containing automobile desing and performance information:

```{r}

head(mtcars)          # View first 6 rows of the dataframe
```

The first row in a dataframe is called the *header* and holds all  column (variable) names. The first column of the dataframe usually consists of rownames. However, rownames aren't always present in dataframes. All rows after the *header* are called *data rows*. Each member of a row is called a *cell*. 

We can find out how many rows and columns are in our dataframe

```{r}

nrow(mtcars)    # Number of rows in mtcars

ncol(mtcars) # Number of columns in mtcars
```

We can also determine the value of a particular cell by specifying the row and column number from which we wish to extract values.

```{r}
mtcars[2,3]   # Value from second row and third column
```

This is also possible using column and row names 

```{r}

mtcars["Mazda RX4 Wag", "disp"]
```

### EXERCISE 1

1. Create the following vectors and merge them to create a dataframe  
- grade = [A, B, A, C, D, B] 
- subject = [Science, Geography, Math, English, History, French]
- credits  = [4,4,3,2,3,3]


2. Determine the number of rows and columns in the dataframe.

***

## <a name = "drop"><a/> Drop or add columns

More often than not, dataframes are not structured in a way that is most practical for our analyses. Handling large dataframes may be complicated, data might lack information that we need, etc. We are able to perform several manipulations to dataframes to make handling data more convenient. 

**Dropping columns:** Large dataframes may be troublesome to handle, and may hold information that is not necessary for our particular analyses. We can easily select columns we wish to delete to make our dataframe more manageable. 

We can modify the dataframe by specifying columns we wish to retain:

```{r}

iris.drop = iris[,2:5] # Create new dataframe with columns 2-10 only

head(iris.drop)
```

Another way to do this is by specifying which column is to be dropped preceded by the *minus* (-) sign as follows:

```{r}

iris.drop = iris[,-1] # New dataframe without the first column 

names(iris.drop)        # Print column names of the new dataframe
```

**Adding columns** : Other times we might want to add information to a dataframe. This is done with the *cbind* function, as with vectors or matrices. 

Let's assume we wish to add an ID number for each row in the dataframe:

```{r}

ID = 1:nrow(iris)           # Create a vector with as many elements as there rows in the dataframe

names(iris)                 # Check column names before adding X

iris.add = cbind(ID, iris)  # Add ID column

names(iris.add)             # Column names after adding ID
```

Keep in mind that the order in which you enter the elements in the *cbind* function will determine where the column is added. If the vector is entered first, then it will be the first column of the new dataframe, otherwise it will be the last column. 

###EXCERCISE 3

1. Drop the following columns from the iris dataframe : Sepal.Width, Petal.Width

2. Add an column with ID number to the mtcars dataframe

***

## <a name = "dsum"><a/>Data summary 

We can take a quick glimpse of our dataframe by using str(). 

```{r}

str(iris)
```

This will let us visualize names, data type and first elements of each column.  


### EXCERCISE 4

1. Summarize your dataframe.

***

## <a name = "colname"><a/>Change column names

Column and row names can be changed by the colnames() and rownames() functions respectively.

```{r}

iris.head = head(iris)        # New data frame with only first 6 rows of iris


colnames(iris.head) = c("sep_len", "sep_wid", "pet_len", "pet_wid", "spp")    # Change column names

rownames(iris.head) = c("ID1", "ID2", "ID3", "ID4", "ID5", "ID6") # Change row names

iris.head #print dataframe
```

You can also change the name of a particular column by specifying its number

```{r}
colnames(iris.head)[2]  = "SepalWidth" # Specifiy the number of the column being renamed

names(iris.head)              # Print new column names
```

***

## <a name = "subset"><a/>Subsetting the dataframe

There may be times when we might want to manipulate smaller sets of the original dataframe. For example we may be interested only in flowers with petals longer than 4mm. We can do this by using mathematical operators

```{r}

long_petal; = iris[iris$Petal.Length > 4,]  # Get all rows with petals longer than 4
                                             # The comma is important here

str(petal)

```

Or we might be interested in a specific species

```{r}

setosa = iris[iris$Species == "setosa",]   # Get all rows with Species = Setosa 
                                        # The comma is important here

head(setosa)                             # Print first values of the dataframe
```

However if we use *str*() we will notice that the other levels for species have been retained. This happens because Species is a *factor*. 

```{r}
str(setosa)
```
To succesfully subset dataframes by factors we must use *drop.levels*(). 

```{r}
setosa = droplevels(iris[iris$Species == "setosa",])
```

Furthermore, we can exclude rows with certain values by using logical operators

```{r}

other.spp = drop.levels(iris[iris$Species != "setosa",]) # Get all rows with species different from setosa
                                            # != means different than
str(other.spp)
```

### EXCERCISE

Using the mtcars dataframe create different subsets for:

1. All cars with  4 cylinders

2. Cars with more than 25 miles per galon (mpg)

3. All gears except 3



