---
title: "Descriptive Statistics"
author: "EcoHealth Alliance"
date: "February 8, 2016"
output: html_document
---

<img src = "rlogo.png" style="position:absolute;top:0px;left:0px;"/>


## Contents

[Quantitative vs qualitative data](#qvq)

[Frequency distribution: qualitative](#frequal)

[Frequency distribution : quantitative](#frequant)

[Plotting frequency: qualitative](#barplot)

[Plotting frequency: quantitative](#hist)

[Plotting frequency: qualitative](#barplot)

[Numerical measures](#numes)

[Quartiles](#quart)

[Mean](#mean)

[Median](#median)

[Variance and standard deviation](#varsd)

[Boxplots](#box)

##<a name = "qvq"><a/> Quantitative vs qualitative data

Data can be separated into two categories depending on how they represent observations or measurements of a given variable. 

Data is *quantitative* if measurements can be represented numerically, in other words if data can be quantified (e.g. age, weight, height).

Furthermore, *quantitative* data can be either *continuous* if it can take any value within a range (e.g. height, weight) or *discrete* if it can only take specific values (e.g. age).

*Qualitative* data on the other hand, refers to non-numerical, nominal data (e.g. eye color, yes-no answers).

It is important to understand the differences between these types of data in order to properly choose the statistical analyses to be applied. For this reason we will first consider methods related to qualitative data and then will move on to methods for the analysis of quantitative data. 

## Frequency distribution

### <a name = "frequal"><a/>Qualitative Data 

Frequency distributions summarize occurence within a set of non-overlapping values. For discrete data we can use the table() function to obtain the frequency distribution. 

```{r, echo = FALSE, cache.comments = TRUE}

movies = read.csv(file = paste(getwd(),"data/movies.csv", sep = "/"), sep = ",", header= T)

```

```{r}

genres = movies$genres     # genres in our movie dataframe

gen.freq = table(genres)   # get frequency distribution

gen.freq                   # print frequency distribution of genres

```

We can then transform this into a column format using cbind().

```{r}

cbind(gen.freq)
```

We can also look at the proportional or relative frequency of data rather than absolute counts.

```{r}

gen.rel.freq = gen.freq/nrow(movies)   # Calculate relative frequency 

cbind(gen.rel.freq)

```

### <a name = "frequant"><a/> Quantitative data

For quantitative data this is more challenging since measurements are continuous and may overlap. For this reason we must break the data into classes to determine frequency of data within these categories. For that we must:

1. Determine the range of our data 

```{r}

ratings = movies$rating    # create vector for ratings of movies

range(ratings)             # range of our data
```

It shows that our data ranges approximately from 6.74 to 9.30.

2. Break the data into even non-overlapping classes. Since we determined that our data ranges from [6.74, 9.30],  we can set the classes the range of [6.5, 9.5] split by 0.5 intervals.

```{r}

breaks = seq(6.5, 9.5, by = 0.5)   # set break sequence by 0.5

breaks
```

3. Fit the data into the intervals we created.

```{r}

ratings.cut = cut(ratings, breaks)  # cut our data according to predetermined breaks  
```

4. Finally, create the frequency table as we do with qualitative data

```{r}

ratings.freq = table(ratings.cut) # sort our data withing break classes

cbind(ratings.freq)
```

As with qualitative data, we can determine the relative frequency.

```{r}
ratings.rel.freq = ratings.freq/nrow(movies) # relative frequency

cbind(ratings.rel.freq)
```


### EXCERCISE 1

Load the tv show dataframe and determine:

1. Overall frequency distribution of days that shows are aired (airs_day)

2. Relative frequency of networks in which shows are aired

***

## Plotting frequency

###<a name = "barplot"><a/> Qualitative data

Once we have created our frequency distribution tables we are able to graphically visualize this information using bar plots.

```{r}

barplot(gen.freq, main = "Movie Genres")  # plot frequency
```

However we notice that axis labels do not fit properly. We can change the graphic parameters to make our plot look better. 

```{r}

barplot (gen.freq, main = "Movie Genres", las =3, cex.names = 0.7) # change font size and position
```

### EXCERCISE 2

Using the tv show dataframe, plot the frequency days in which shows are aired.

###<a name = "hist"><a/> Quantitative data

To plot quantitative data however, it is not necessary to construct a frequency table first. We can visualize frequency by creating a histogram using our data directly.

```{r}

hist(ratings, col = "gray")   # histogram of ratings
```

### EXCERCISE 3

Using the tv show dataframe create a histrogram for aired episodes and year.

***
 
## <a name = "numes"><a/>Numerical measures

Even though frequency gives us an idea of how our data is distributed, there are several numerical measures commonly used in statistics that help us further understand the distribution of our data. We will now briefly explore quartiles, measures of central tendency and measures of variation.

###<a name = "quart"><a/> Quartiles

Quartiles divide our data into four equal parts and allow us to know the proportion of our data that falls under a certain value. The first or lower quartile cuts off the first 25% of the data, the second quartile (or median) cuts at 50%, and the upper (third) quartile at 75%.


```{r}

quantile(ratings)
```


We can also see how much the middle portion of our data spreads by calculating the *interquantile range*

```{r}
IQR(ratings)
```

### EXCERCISE 4

Determine the quartiles and interquartile range for number of aired episodes in the tv show dataframe.

###<a name = "mean"><a/> Mean

The mean or arithmetic average represents the sum of all values of our data divided by the number of observations in our data. We can determine the mean 

```{r}

mean(ratings)
```

### <a name = "median"><a/>Median

The median is a value that separates data in half. Meaning that half of the values in the data are higher than the median while the other half is lower. 

```{r}

median(ratings) 
```

Note that when our data follows a normal distribution, which just so happens to be the case with ratings, the mean and the median will be very similar.

However, when data is skewed these will differ. This is because the mean is greatly affected by extreme values. 

### <a name = "varsd"><a/>Variance and standard deviation

*Variance* is a measure of how much our data is dispersed around the mean. In other words, it is the average distances of our sample values from the mean. This helps us understand how much variation is there within our sample population. 

```{r}
var(ratings)
```

*Standard deviation* is the square root of the variance. Many times it is prefered to represent variation since it usually yields more manageable values than the variance. However, both values are equally valid and may be used to asses variation within a particular population. 

```{r}
sd(ratings)
```

### EXCERCISE 5 

Determine the mean, median, variance, and standard deviation for *run time*, *aired episodes* and *year* of the tv show dataframe.
***

## <a name = "box"><a/>Box Plots

We can graphically represent how our data is distributed by creating box plots. These will allow us to visualize quartiles, the mean, and general variation of our data. 

```{r}

boxplot(ratings, horizontal = TRUE)
```

### EXCERCISE 6

Create a boxplot to describe the number of aired episodes for tv shows.


