---
title: "Introduction to R | Statistical Test examples"
author: "EcoHealth Alliance"
output: github_document
---
<img src = "rlogo.png" style="position:absolute;top:0px;left:0px;"/>


## Contents

[T-test](#ttest)

[Anova](#anov)

[Chi square](#chi)

[Linear regression](#lm)



R can be used to perform a wide range of powerful statistical analyses ranging from parametric statistics to spatial analyses and more. In this section we will display just a few examples of statistical tests you can perform with this software.

We will be using the *Fatalities* dataset from the *AER* package to show several statistical applications of **R**. This dataframe contains annual US traffic fatalities of the "lower 48" states from 1982-1988. 

```{r}
library(AER)       #load library and dataframe
data("Fatalities") #Remember to install the package if you do not have it
```

### <a name = "ttest"><a/> T-test

To perform T-test we can use the ```t.test()``` function. For example, let's evaluate if there are any significant differences in number of fatalities between driviers aged 15-17 and 21-24 years old

```{r}
tt1 = t.test(Fatalities$fatal1517, Fatalities$fatal2124) 
```

Since `p = 1.995e-15` we observe significant differences in number of fatalities between the different age groups. 

Let's suppose that we would now like to see if there are any differences between locations where alcohol breath tests are applied and where these are not applied.

```{r}
tt2 = t.test(Fatalities$fatal~ Fatalities$breath)
```  

As we can see, there seems to be a significantly lower number of fatalities in locations where alcohol breath tests are required by law. 

### <a name = "anov"><a/> ANOVA

R is also capable of performing ANOVA tests. Let's apply a single factor ANOVA for number of fatalities per state in the year 1988. 

```{r}
fat88 = droplevels(Fatalities[Fatalities$year == 1988,])

anv1= aov(Fatalities$fatal ~ Fatalities$state)
```

### <a name = "lm"><a/> Linear regression


```{r}
lm = lm(fatal ~ youngdrivers + pop + drinkage, data = Fatalities)