---
title: "Read spss files"
author: "EcoHealth Alliance"
date: "March 2, 2016"
output: html_document
---

<img src = "rlogo.png" style="position:absolute;top:0px;left:0px;"/>

### Method 1 (recommended)

This is the most straightforward and easiest way to read SPSS files into R. 

**Step 1:** Load the *foreign* package

```{r}
library(foreign) # load foreign R package
```

**Step 2:** Now we proceed to upload the SPSS file with read.spss()

```{r eval = FALSE}
data = read.spss(file.choose(), to.data.frame = TRUE)
```

Voila! you're all set. 

Note that setting to.data.frame as TRUE is what actually allows R to treat our data as an R dataframe. 

### Method 2

There might be cases in which method 1 does not work. In that case do not despair, there is another way to load your SPSS files onto R though it may be a little more complicated.

**Step 1:** First we need to install the *mesmic* package (if you have not done so already) and load it. 

```{r eval = FALSE}

install.packages("memisc")

library(memisc)
```

**Step 2:** Then we read the file as an R data set

```{r eval = FALSE}

data= as.data.set(spss.system.file(file.choose()))
```

**Step 3:** Transform the data set object into an R dataframe

```{r eval = FALSE}
data = as.data.frame(data)
```

You should now be able to work with your SPSS dataframe on R.
